---
title: "Homework 4"
author: "Chengyi Wang"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
---

## Instructions
Answer the following questions and/or complete the exercises in RMarkdown. Please embed all of your code and push the final work to your repository. Your report should be organized, clean, and run free from errors. Remember, you must remove the `#` for any included code chunks to run. Any plots must have appropriate titles and aesthetics.  

## Load the tidyverse
```{r}
library("tidyverse")
options(scipen = 999) #disable scientific notation
```

For this homework, we will be using the `msleep` dataset from the `ggplot2` package. This dataset contains information about the sleep habits of various mammals. [Reference](https://ggplot2.tidyverse.org/reference/msleep.html): V. M. Savage and G. B. West. A quantitative, theoretical framework for understanding mammalian sleep. Proceedings of the National Academy of Sciences, 104 (3):1051-1056, 2007.

**1. Use a summary function of your choice to get an idea of the structure of the msleep data.**
```{r}
glimpse(msleep)
```

**2. What are the names of the variables? To get a better idea of what the variables mean, you can use `?msleep`.**
```{r}
names(msleep)
?msleep
```

**3. Make a plot that shows the number of mammals in each vore type. Which vore type is most represented in the data?**
```{r}
ggplot(
  data = msleep, mapping = aes(x = vore, fill = vore)
  )+
  geom_bar()+
  labs(title="Number of Each 'vore' Type", x="vore type")
```

**4. What is the average length of time that each mammal sleeps?**
```{r}
mean(msleep$sleep_total)
```

**5. Use `summary()` to get a quick idea of the range of values for the variable `bodywt`.**
```{r}
summary(msleep$bodywt)
```

**6. Build a plot that shows the range of `bodywt.` by vore type. Because the range of `bodywt` is so variable, you will need to scale the y-axis. You do this by adding `scale_y_log10()` as another layer.**
```{r}
ggplot(
  data = msleep, mapping = aes(x = vore, y = bodywt, colour = vore)
  )+
  geom_boxplot()+
  scale_y_log10()+
  labs(title = "Body Weight Range of Each 'vore' Type", x="vore type", y="Logarithmic Scale Body Weight (kg)")
```

**7. Is there a relationship between body weight and total sleep time? Add appropriate labels and a title to your plot.**

There is relationship, but not very strong. 
```{r}
ggplot(
  data = msleep, mappin = aes(x=bodywt,y=sleep_total)
  )+
  scale_x_log10()+
  geom_point()+
  geom_smooth(method=lm)+
  labs(title = "Body Weight vs. Total Sleep Time", x="Logarithmic Scale Body Weight (kg)", y="Total Sleep Time")
```

**8. Which vore type tends to sleep the most? Build a plot to support your answer.**

Insectivore tends to sleep the most. 
```{r}
ggplot(
  data = msleep, mapping = aes(x = vore, y = sleep_total, colour = vore)
  )+
  geom_boxplot()+
  labs(title = "Total Sleep Time of Each 'vore' Type", x="vore type", y="Total Sleep Time (hours)")
```

**9. Is there a relationship between brain weight and body weight? Build a plot to support your answer.**

Yes, brain weight and body weight are positively correlated. 
```{r}
ggplot(
  data = msleep, mappin = aes(x=bodywt,y=brainwt)
  )+
  scale_x_log10()+
  scale_y_log10()+
  geom_point()+
  geom_smooth(method=lm)+
  labs(title = "Body Weight vs. Brain Weight", x="Logarithmic Scale Body Weight (kg)", y="Logarithmic Scale Brain Weight (kg)")
```

**10. Build one plot of your choice that provides additional insight into the msleep dataset. Be sure to include appropriate labels and a title.**

Total sleep time and REM sleep time are correlated. 
```{r}
ggplot(
  data = msleep, mappin = aes(x=sleep_total,y=sleep_rem)
  )+
  geom_point()+
  geom_smooth(method=lm)+
  labs(title = "Total Sleep Time vs. REM Sleep Time", x="Total Sleep Time (hours)", y="REM Sleep Time (hours)")
```

## Knit and Upload
Please knit your work as an .html file and upload to Canvas. Homework is due before the start of the next lab. No late work is accepted. Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  